---
title: "TØ 3.1"
output: html_document
date: "2024-09-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Copying and updating stopping criteria in Asger's code
```{r}
##-----------------------------------------------------
## Earth quakes data
##-----------------------------------------------------
## Number of major world-wide earth quakes between 1900 and 2006 
ObsSeq <- c(13,14,8,10,16,26,32,27,18,32,36,24,22,23,22,18,25,21,
            21,14,8,11,14,23,18,17,19,20,22,19,13,26,13,14,22,24,
            21,22,26,21,23,24,27,41,31,27,35,26,28,36,39,21,17,22,
            17,19,15,34,10,15,22,18,15,20,15,22,19,16,30,27,29,23,
            20,16,21,21,25,16,18,15,18,14,10,15,8,15,6,11,8,7,18,
            16,13,12,13,20,15,16,12,18,15,16,13,15,16,11,11)
##-------------------------------------------------------
## EM algorithm for two component Poisson mixture model
##------------------------------------------------------
## Starting values
alphaEst <- 0.6
lamEst <- 2
etaEst <- 4
logLk <- sum(log(alphaEst*dpois(ObsSeq,lamEst)+
                     (1-alphaEst)*dpois(ObsSeq,etaEst)))
## Number of iterations
nIter <- 10
eps <- 10^-9 #Stopping criteria
## Run EM algorithm
iter <- 1
while(T){
  xPrb <- alphaEst*dpois(ObsSeq,lamEst)/
    (alphaEst*dpois(ObsSeq,lamEst)+
    (1-alphaEst)*dpois(ObsSeq,etaEst))
  alphaEst <- sum(xPrb)/length(ObsSeq)
  lamEst <- sum(xPrb*ObsSeq)/sum(xPrb)
  etaEst <- sum((1-xPrb)*ObsSeq)/sum(1-xPrb)
  if(abs(logLk - sum(log(alphaEst*dpois(ObsSeq,lamEst)+
                     (1-alphaEst)*dpois(ObsSeq,etaEst)))) < eps) #Stop if updates are "short"
  {
    break
  }
  
  logLk <- sum(log(alphaEst*dpois(ObsSeq,lamEst)+
                     (1-alphaEst)*dpois(ObsSeq,etaEst))) #Otherwise update value of logLk
  cat("Iteration:",iter,
      "alpha:",alphaEst,"lam:",lamEst,"eta:",etaEst,
      "logLk",logLk,"\n")
  iter <- iter + 1
  
  }
##------------------------------------------------------
plot(ObsSeq,type="l")
points(1:length(xPrb),5+10*(1-xPrb),col="red",type="l")
##------------------------------------------------------
mx <- max(ObsSeq)
plot(1:mx,tabulate(ObsSeq),type="l",lwd=2)
points(1:mx,length(ObsSeq)*alphaEst*dpois(1:mx,lam=lamEst),
       col="red",type="l",lwd=2)
points(1:mx,length(ObsSeq)*(1-alphaEst)*dpois(1:mx,lam=etaEst),col="blue",
       type="l",lwd=2)
points(1:mx,
       length(ObsSeq)*(alphaEst*dpois(1:mx,lam=lamEst)+
                      (1-alphaEst)*dpois(1:mx,lam=etaEst)),
       col="green",type="l",lwd=2)

```

c) Values in the book are
alpha = 0.676, 0.425
lambda = 15.777, 26.840,
logLk = 360.369 (we get the same :-))

d) The m = 1 case

```{r}
lambda <- mean(ObsSeq)
negLogLk <- -log(prod(dpois(ObsSeq, lambda)))

c(lambda, negLogLk)
```
We get the same numbers :)

e,f) Done by hand

g)
The function below calculates the log likelihood for any given poisson mixture
(can handle any given number of components)
```{r}
log_like <- function(deltas, lambdas, data = ObsSeq)
{
  m <- length(deltas)
  n <- length(data)
  temp <- numeric(n)
  for(i in 1:m)
  {
    temp <- temp + deltas[i] * dpois(data, lambdas[i]) #This loop calculates for each observation how much each component adds to the log likelihood
  }
  return(sum(log(temp)))
}
```

we make functions to update the delta values and lambda values based on what we
derived in the E and M step.

```{r}
update_delta <- function(deltas, lambdas, data = ObsSeq)
{
  n <- length(data)
  m <- length(deltas)
  prob_mat <- outer(ObsSeq, lambdas, dpois) #Calculates P(Y_i = y_i | X_i = j) for all j = 1,...,m
  X_probs <- c()
  for(i in 1:m)
  {
    X_probs <- cbind(X_probs, prob_mat[,i]*deltas[i])
  }
  return(colSums(X_probs / rowSums(X_probs)) / n)
}

update_lambda <- function(deltas, lambdas, data = ObsSeq)
{
  n <- length(data)
  m <- length(deltas)
  prob_mat <- outer(ObsSeq, lambdas, dpois)
  X_probs <- c()
  for(j in 1:m)
  {
    X_probs <- cbind(X_probs, prob_mat[,j]*deltas[j])
  }
  Enjs <- colSums(X_probs / rowSums(X_probs))
  Esjs <- colSums(ObsSeq * X_probs / rowSums(X_probs))
  return(Esjs/Enjs)
}
```

Notice that I have chosen to return everything in a tabular form - however this
has been done a little lazily so that all values except for the lambdas and deltas
are actually just repeated to make the table work.

```{r}
EM_run <- function(deltas, lambdas, eps, data = ObsSeq)
{
  iter <- 1
  log_likes <- c(log_like(deltas, lambdas, data))
  while(T)
  {
    deltas_now <- deltas 
    lambdas_now <- lambdas #if we don't do this we update lambdas using updated deltas
    iter <- iter + 1
    deltas <- update_delta(deltas_now, lambdas_now, data)
    lambdas <- update_lambda(deltas_now, lambdas_now, data)
    logLk <- log_like(deltas, lambdas, data)
    log_likes <- c(log_likes, logLk)
    #cat("iteration:", iter - 1, "deltas:", deltas, "lambdas:", lambdas,
        #"loglK:", loglK)
    if(abs((log_likes[iter]-log_likes[(iter - 1)])/log_likes[(iter - 1)]) < eps)
    {
      break
    }
  }
  m <- length(deltas)
  logLK <- rep(-logLk, m)
  p <- m + m-1 #number of parameters, notice that we only have m-1 free parameters from delta since delta_j = sum_{i =/= j} delta_i
  AIC <- rep(2*(-logLk + p),m)
  BIC <- rep(-2*logLk + log(length(ObsSeq)) * p,m)
  EX <- rep(sum(deltas*lambdas),m)
  var_val_temp <- 0
  for(i in 1:m)
  {
    for(j in 1:m)
    {
      if(i < j)
      {
        var_val_temp <- var_val_temp + deltas[i]*deltas[j]*(lambdas[i]-lambdas[j])^2 #From exercise 1.1 b
      }
    }
  }
  VarX <- EX + var_val_temp
  
  return(round(cbind(deltas, lambdas, logLK, AIC, BIC, EX, VarX),3))
}
```

h)

```{r}
#repeat case 0: m = 2
deltas <- c(0.6, 0.4)
lambdas <- c(2, 4)
eps <- 10^-15
EM_run(deltas, lambdas, eps)
```

```{r}
#case 1: m = 3
deltas <- c(1/6, 1/3, 1/2)
lambdas <- c(10, 20, 30)
eps <- 10^-15
EM_run(deltas, lambdas, eps)
```

#Play around with the code above to get even more components! :)

#Figure 1.4
```{r}
hist(ObsSeq, freq = F, breaks = (0:14)*3, ylim = c(0, 0.1), ylab = "", xlab = "",
     main = "m=3")
lines(0:50, sapply(0:50, function(x){0.278*dpois(x, 12.736) + 
    0.593*dpois(x, 19.786)+0.130*dpois(x, 31.630)}))
```

i)
Choice is between m = 2 and m = 3. We consider the empirical variance of the
data as well as AIC/BIC and plots

```{r}
var(ObsSeq) #close to the variance we find for m = 3
```
Based on this and all of the above we choose m = 3

j)
```{r}
acf(ObsSeq)
```
#Does not look independent.

#Compare to data generated from the fitted model (under independence)
```{r}
deltas <- c(0.278, 0.593, 0.130)
lambdas <- c(12.736, 19.786, 31.630)
n <- length(ObsSeq)
simDat <- numeric(n)
for(i in 1:n)
{
  Z <- sample(1:3, 1, prob = deltas)
  simDat[i] <- rpois(1, lambdas[Z])
}

acf(simDat)
```



